---
title: GBIF marine datasets
date: "`r Sys.Date()`"
author: Pieter Provoost
output: (function(...) {
  rmdformats::robobook(toc_depth = 3, ...) })
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })  
---

This notebook uses the GBIF API to generate statistics on marine data in GBIF. Note that this is a preliminary analysis which only uses species. The process looks like this:

1. Get all marine species names from WoRMS
2. Using the species names from WoRMS, get all profiles of marine species from GBIF
3. Using the `nubKey`s from the species profiles, get record counts by dataset from GBIF 

## Get all marine species from WoRMS

This uses a full export of the WoRMS database. This file is not included in the repository, contact the WoRMS team to get access to such an export.

```{r message=FALSE, warning=FALSE, echo=FALSE}
library(dplyr)
library(data.table)

taxon <-  fread("worms_export/taxon.txt", sep = "\t", na.strings = "", quote = "") %>%
  as_tibble()

speciesprofile <-  fread("worms_export/speciesprofile.txt", sep = "\t", na.strings = "", quote = "") %>%
  as_tibble()

taxon_marine <- taxon %>%
  left_join(speciesprofile, by = "taxonID", suffix = c("", ".y")) %>%
  filter(taxonomicStatus == "accepted" & isMarine == 1 & taxonRank == "Species")

species <- unique(taxon_marine$scientificName)
```

## Fetch species profiles from GBIF

Here we use the GBIF API to search for species by name. The results are stored as CSV files in the `profiles` folder.

```{r message=FALSE, warning=FALSE}
library(stringr)
library(jsonlite)
library(progress)

pb <- progress_bar$new(total = length(species), format = "[:bar] :current/:total (:percent) ETA: :eta")

for (sp in species) {
  key <- str_replace(tolower(sp), "\\s", "_")
  filename <- paste0("profiles/", key, ".csv")
  if (!file.exists(filename)) {
    url <- URLencode(paste0("https://api.gbif.org/v1/species?name=", sp))
    res <- fromJSON(url)$results
    if (length(res) > 0 & "nubKey" %in% names(res)) {
      species_names <- res %>%
        select(key, nubKey, nameKey, taxonID)
      write.csv(species_names, filename, row.names = FALSE, na = "")
    } else {
      write.csv(data.frame(nubKey = character(0)), filename, row.names = FALSE, na = "")
    }
  }  
  pb$tick()
}
```

## Get occurrence counts by dataset for each species

Here we use another API endpoint to get the number of records per dataset for each species. But first read all the CSV files generated in the previous section:

```{r message=FALSE, warning=FALSE}
library(purrr)

files <- list.files(path = "profiles", pattern = "*.csv", full.names = TRUE)

profiles <- map(files, ~read.csv(.)) %>%
  bind_rows()
```

Now let's get the dataset statistics. Results are stored in the `statistics` folder as CSV files.

```{r message=FALSE, warning=FALSE}
nubkeys <- na.omit(unique(profiles$nubKey))

pb <- progress_bar$new(total = length(nubkeys), format = "[:bar] :current/:total (:percent) ETA: :eta")

for (nubkey in nubkeys) {
  filename <- paste0("statistics/", nubkey, ".csv")
  if (!file.exists(filename)) {
    url <- URLencode(paste0("https://api.gbif.org/v1/occurrence/counts/datasets?nubKey=", nubkey))
    res <- fromJSON(url)
    if (length(res) > 0) {
      df <- data.frame(dataset = names(res), records = unlist(res))
      write.csv(df, filename, row.names = FALSE, na = "")
    } else {
      write.csv(data.frame(dataset = character(0), records = integer(0)), filename, row.names = FALSE, na = "")
    }
  }  
  pb$tick()
}
```

Once all statistics have been fetched, read them from CSV:

```{r message=FALSE, warning=FALSE}
files <- list.files(path = "statistics", pattern = "*.csv", full.names = TRUE)

statistics <- map(files, ~read.csv(., colClasses = c("character", "integer"))) %>%
  bind_rows()
```

## Calculate statistics

```{r message=FALSE, warning=FALSE}
stats <- statistics %>%
  group_by(dataset) %>%
  summarize(records = sum(records)) %>%
  arrange(desc(records))

n_datasets <- format(nrow(stats), big.mark = ",")
n_records <- format(sum(stats$records), big.mark = ",")

stats %>%
  rmarkdown::paged_table()
```

In total we have found `r n_datasets` datasets containing marine species, for a total of `r n_records` marine species records.
